PLANETESIMAL ACCRETION AT SHORT ORBITAL PERIODS

### 2023-04-03 refereeResponse.txt

Specific comments, notes and suggestions.

TODO: First few edits need to be highlighted in manuscript with strikethroughs and bold face.

Abstract
7: Are short period planets that exotic in the broader context of exoplanets?
	Hard to say! We have replaced 'exotic' with 'wide variety'.
14: "[...] population OF dynamically [...]"
	Fixed
17-18: something is missing in the second part of this sentence.
	Missing words at the end of L17: 'investigate *the effects*'

1.Introduction
26: How large are these solids? I assume this is referencing micron sized dust-grains?
	Correct, some text was added to clarify this.
68: Add reference
	Added a handful of references arguing that the asteroids, KBOs and Oort cloud bodies are leftover planetesimals.
	See Duncan, Tremaine Quinn papers (particularly 1988 one + references?)
72: Maybe elaborate on this. How long is the period of quiescence? How does this instability work?
	We intentionally left this part vague as someone really needs to look into this more! There are three relevant timescales here:
	1. The timescale for the instability itself (depends on the orbital properties of the embryos, see Chambers 98)
	2. Dynamical friction between protoplanets and planetesimals. This will shut off once the planetesimals are either scattered away, or slowly accreted onto the embryos. Not sure what the timescale for this would be. Note that this shouldn't operate at short period according to this work!
	3. Tidal eccentricity damping of the embryos due to the gas disk. Should shut off once the gas disk disperses on ~Myr timescales (Beckwith + Sargent 1996).
	TODO: decide what to add to the text here.
88: Add reference maybe.
	I don't know of any references which have tested the applicability of statistical methods during the oligarchic growth phase, but maybe some clarification would help. The runaway growth phase produces a small number of large bodies, which eventually interact and terminate runaway growth (Ida + Makino 1993, Kokubo + Ida 1995). Given how few oligarchs are produced, one would not expect particle-in-a-box methods to be a good approach here. We've clarified this in the text as well.
99: Capitalize "Systems of Tightly-packed Inner Planets"
	Fixed
104: Gas-Disk driven migration → reference?
	Added reference to Izidoro 17 and 21, which invokes gas disk driven migration, followed by resonant trapping and an possible orbital instability to explain the observed architectures of STIPs
116: What is the "isolation mass" initial condition? Maybe isolation mass can be defined
	We will define isolation mass later in the paper. For now, we removed this term and replaced it with 'fully-formed, evenly spaced protoplanets'.
118: I'm not sure if this belongs in the introduction. I think the mention of future work at the end of the discussion is sufficient.
	Agreed. Removed the last sentence of this paragraph.

@@

2.2 Planetesimal Size and Extent of Hill Sphere
253: "will remain proportional" - Proportional to what?
	Changed the wording to 'ratio between the relaxation and collision timescale remains constant' to clarify this.
254: but this doesn't have to be the case at all orbital periods, right? Given Fig.1 it seems clear that outside of 1AU, and especially for the Kuiper Belt this condition seems to be fulfilled.
	Exactly! That's the point we're getting at here. (TODO: do I need to clarify something in the paper here?)
263: Again, I think the concept of Hill radius and tidal stability needs to be formally introduced earlier.
	TODO: Introduce this formally earlier
273: Wouldn't these bodies then be tidally disrupted, and therefore not exist? Alpha > 1 is certainly not true for any solar system bodies that I am aware of. I understand that the scaling of alpha predicts alpha >1 at some radius (which I would explicitly mention), but this just implies that inside this radius, bodies must have greater densities to not be disrupted.
	The distance at which alpha = 1 (a_alpha) and the Roche (a_roche) limit are similar, but not identical. For a rigid spherical body, a_roche ~ 0.88 a_alpha. Applying the hydrostatic correction, a_roche ~ 0.6 a_alpha. This is important, because it implies that the new growth mode presented in this paper should be relevant for ring systems! We will add a note of this in the text. I think some of the confusion comes from the fact that the Roche limit and the Hill sphere size are not derived using the same formalism.
289: gravitational radius: is this physically motivated? Maybe reference
	There is a physical explanation of r_g immediately below equation 5 in the text. Also added a reference to Ida 1990.
296: beta > 1 means collision cross section > geometric and small gravitational focusing. I recommend explicitly stating what the physical intuition for beta < 1 is.
	Done.

### 2023-04-04 refereeResponse.txt

5.2. Gas Drag Force
512: I think it would be very informative to calculate the dimensionless stopping time, i.e. the Stokes number of the planetesimals, I'm assuming that they are substantially above unity, which means that the bodies are essentially decoupled from the gas.
	Agreed. Describing things in terms of the Stokes number allows us to simplify this section a bit. For the initial planetesimals, St >> 1 at both the inner and outer disk edge Because St ~ m^1/3, we can guarantee that bodies will remain decoupled from the gas throughout the simulation. By discussing the Stokes number, we can also avoid having to compare the stopping time with the integration time of the simulations. These points are now all discussed in the text.
536: Isolation mass lacks definition
	We have removed any mention of the isolation mass here. The fact that St ~ m^1/3 is enough to guarantee that the gas and solids remain decoupled.

5.3 Time-stepping Criterion
554: footnote 2 can just go in the main text body.
	Done
584: footnote 3 can also go in the main text body.
	Done
584: Are there any works on the numerical robustness of this more efficient time-stepping scheme. While I'm inclined to believe that the error is only marginal, I think having some sort of test (or a reference to one) would make the results more steadfast.
	I'm fairly certain that we are the first ones to try this. I've added an appendix that compares this timestepping scheme against a fixed timestep case. There appear to be no meaningful differences. TODO: Show mass distribution as differential or cumulative?

5.4 Results
586: Maybe include reference to figure where this is shown
	Figure 1 of Kokbubo 2002 shows this clearly. A reference to the paper has been added in the text.
592: Is this really always true? Counterexample e.g. Emsenhuber et al 2021 and others.
	The NGPPS models (along with the Clement 2020 paper) seem to be a couple of notable exceptions to this, although the embryo growth is not treated self-consistently leading up to the giant impact phase. We have added a mention of the Emsenhuber papers to the text.
Table 2:
What is the rescaled integration time?
	TODO: Not sure where to explain table columns now?

Figure 5:
I think the bottom panel should have a log y-axis
I would make the color of the dots consistent with Fig.2, where the final outcome is orange
There could be numbers on the contours indicating the density of the initial condition.
The meaning of the dashed line in the bottom panel could be indicated inside the figure itself.
I'm confused by the statement "interior to the oligarchic growth region (around 60d)". Where does this definition come from? Is this defined by bodies exceeding the critical feeding size? If so the statement should be rephrased to make the causal connection clear.

606: How can a mass be expressed in units of feeding zone size?
620: What is the intuition behind tilde b, how exactly is it defined? Since it is a dimensionless quantity, calling it an orbital separation is not intuitive to me. Can tilde b be converted to a physical size? Also, I struggle to interpret the critical tilde be of 2*sqrt(3). Do I understand correctly that a body with tilde b less than 2 sqrt(3) is associated with a feeding zone so small that close encounters are not possible, and thus are not expected to grow further?
632: Why are the bodies expected to follow the tilde b = 2sqrt(3) from a theoretical perspective? Also, I'm unconvinced that the bodies do that at all in the presented simulations - they fall below this value but don't follow it.
640: If anything the opposite seems to be true, no? Maybe have a specific value for the expected eccentricity of protoplanets with and without inelastic damping.
644: If I look at the contour map by eye, the eccentricity distribution doesn't seem all that different in the inner vs outer disk. Maybe the difference can be quantified?
647: This is an interesting point. I'd be interested in seeing a comparison between inner and outer disk accounting for that difference in time scales. Is it possible to include such a comparison maybe with a figure?
Figure 6:
This figure could be improved if each line was colored by the semi-major axis of the radius bin, and if there was a corresponding colorbar. The difference between the 60 day boundary could still be indicated with an additional line.
665: I struggle to see that there is a significant qualitative difference in the behavior in the lines just inside and just outside the boundary of 60 orbits. The first dashed line also seems to converge to a sigma/Sigma \sim 5 %, not much below the first solid line.
668: add reference. Also footnote 5 can probably go in the main text. Can the authors provide an equation that converts tilde b to units of r_h?
682: This statement seems only to be true for fdshallow. The others have a spread in eccentricity that covers two orders of magnitude...
685: It is not clear to me why more inelastic collisions lead to a more sharply peaked eccentricity distribution, and not just more damping like the authors argue for the fdLo simulation.
701: Again, I don't see that the value for tilde b approaches 2sqrt(3) beyond P=60d orbits. The bodies have smaller feeding zones than tilde b = 2sqrt(3)
702: Are they that similar? Maybe this statement can be more quantitative? To me it looks like the values fall below 2sqrt(3) at 40 days for fdHiShallow and at 70 days for fdLo. The interpretation that the boundary is only driven by the local value for alpha seems somewhat speculative. I'd also be interested how randomness affects this, i.e. if the simulation is initialized with a different random seed, by how much does the final distribution change?

### 2023-04-04 main.tex
Address comment about roche limit
More referee edits in section 5

An insight into the expected behavior in this regime can be gained by
defining a dimensionless parameter, $\alpha$, which is the ratio
between the physical size of a body and its Hill radius, $r_{h}$
\begin{equation}\label{eq:alpha}
	\alpha = \frac{r_{pl}}{r_{h}} = \frac{1}{a} \left( \frac{9 M_{\star}}{4 \pi \rho_{pl}} \right)^{1/3},
\end{equation}
where $a$ is the semimajor axis of the body and $\rho_{pl}$ is its
bulk density. Assuming that the bulk density stays constant as bodies collide and
grow, and that no large-scale migration occurs, the scaling of both
$r_{pl}$ and $r_{h}$ as $m_{pl}^{1/3}$ means that $\alpha$ will be
constant with time. For a composition of ice and rock, $\alpha$ is
small for any presently populated region of the solar system ($\alpha \sim
10^{-2}$ near Earth and $\alpha \sim 10^{-4}$ in the Kuiper belt). As
one moves closer to the sun, $\alpha$ becomes larger than 1, which
implies that the Hill sphere of a body becomes smaller than its physical size.

@@

In addition to the mutual gravitational forces, a Stokes drag force due to the the gas disk is applied to each particle, following the prescription described in section 2.2.1 of \citet{morishima10}. For the initial mass planetesimals, the stopping time (see equation \ref{eq:ts_stokes}) at the inner edge of the disk is roughly 100 yr. Although this is a few orders of magnitude shorter than the integration time of the simulations, the Stokes drag force plays a fairly insignificant role as the bodies grow via mutual collisions. By the time the bodies reach isolation mass (roughly a tenth of an Earth mass), the stopping time at the inner disk edge exceeds 50,000 years. Although the aerodynamic gas drag is not expected to significantly alter the final protoplanet distribution, we include its effects here to be self-consistent with the initial conditions, which were constructed by balancing the effects of viscous stirring with gas drag.

@@

In the case of the fdHi simulation, there are nearly 1 million
particles, whose orbital periods vary by two orders of magnitude. Because the
interaction timescales near the inner edge of the
disk are exceedingly short, a fixed timestep size would required a prohibitively large
number of steps to follow planetesimal growth throughout the entire
disk. For this reason, we use a multi-tiered timestepping scheme, in
which particles are placed onto the nearest power of two timestep
based on their most recently calculated gravitational
acceleration.\footnote{This scheme is used on almost all works using ChaNGa,
and is common among large-scale simulation codes.}

@@

A second issue is introduced when two particles on different timesteps
undergo a collision. As in the previous case, momentum is not
completely conserved because the most recent `kick' steps did not
happen simultaneously for these bodies. Early in the simulation, we
find that this problem tends to trigger runaway growth at the timestep
boundaries first. This issue carries itself forward through the embryo
formation phase, and protoplanets tend to form at the boundaries. To
correct this issue, we ignore collisions between bodies on different
timesteps early in the simulation. We find that preventing
multi-timestep collisions until after the maximum mass grows by a
factor of 10 prevents any artifacts from developing at the timestep
boundaries, while also minimizing the number of `skipped' collisions.\footnote{In the case of fdHi, only about 20 collisions out of an eventual 900,000 are ignored.}

@@

The timescales for embryo formation depend on
the chosen surface density profile, along with the local orbital
timescale. Protoplanets form first at the inner edge of the disk,
where the dynamical timescales are short. Growth proceeds in an
inside-out fashion, with the outermost regions of the disk completing
the protoplanet assembly phase last. This radial timescale dependence is not typically
accounted for in planet formation simulations\footnote{Instead, the innermost protoplanets 
patiently `wait' for this phase of evolution to complete everywhere in the disk}, and appears to be an
important component to forming realistic solar system analogs
\citep{clement20}. As with the narrow annulus simulations, we stop the
integration once the masses of protoplanets in the outermost region of
the disk reach a steady value. In table \ref{tab:sim_properties}, we
summarize the outcomes of the four ``full disk'' cases.

>>>>

An insight into the expected behavior in this regime can be gained by
defining a dimensionless parameter, $\alpha$, which is the ratio
between the physical size of a body and its Hill radius, $r_{h}$
\begin{equation}\label{eq:alpha}
	\alpha = \frac{r_{pl}}{r_{h}} = \frac{1}{a} \left( \frac{9 M_{\star}}{4 \pi \rho_{pl}} \right)^{1/3},
\end{equation}
where $a$ is the semimajor axis of the body and $\rho_{pl}$ is its
bulk density. Assuming that the bulk density stays constant as bodies collide and
grow, and that no large-scale migration occurs, the scaling of both
$r_{pl}$ and $r_{h}$ as $m_{pl}^{1/3}$ means that $\alpha$ will be
constant with time. For a composition of ice and rock, $\alpha$ is
small for any presently populated region of the solar system ($\alpha \sim
10^{-2}$ near Earth and $\alpha \sim 10^{-4}$ in the Kuiper belt). As
one moves closer to the sun, $\alpha$ becomes larger than 1, which
implies that the Hill sphere of a body becomes smaller than its physical size.
\textbf{As an additional note, the Roche limit of the central body and the distance at which $\alpha = 1$ are
similar, but not identical. For a rigid spherical body, $a_{Roche} \simeq 0.88 a_{\alpha = 1}$. Applying a hydrostatic correction to the Roche limit, $a_{Roche} \simeq 0.6 a_{\alpha = 1}$. This accretion mode should
therefore be relevant for ring systems.}

@@

In addition to the mutual gravitational forces, a Stokes drag force due to the the gas disk is applied to each particle, following the prescription described in section 2.2.1 of \citet{morishima10}. For the initial mass planetesimals, the \textbf{Stokes number ($\textrm{St} = t_{s} \Omega$) at the inner and outer disk edge is roughly $2 \times 10^{5}$ and $10^{7}$, respectively. For $\tau_{s} \gg 1$, bodies are decoupled from the gas and are only weakly affected by it. Because $\tau_{s} \sim m^{1/3}$, the Stokes number grows as planetesimal accretion proceeds, and the drag force plays an increasingly minor role.} Although the aerodynamic gas drag is not expected to significantly alter the final protoplanet distribution, we include its effects here to be self-consistent with the initial conditions, which were constructed by balancing the effects of viscous stirring with gas drag.

@@

In the case of the fdHi simulation, there are nearly 1 million
particles, whose orbital periods vary by two orders of magnitude. Because the
interaction timescales near the inner edge of the
disk are exceedingly short, a fixed timestep size would required a prohibitively large
number of steps to follow planetesimal growth throughout the entire
disk. For this reason, we use a multi-tiered timestepping scheme, in
which particles are placed onto the nearest power of two timestep
based on their most recently calculated gravitational
acceleration. \textbf{This scheme is used on almost all works using ChaNGa,
and is common among large-scale simulation codes.}

@@

A second issue is introduced when two particles on different timesteps
undergo a collision. As in the previous case, momentum is not
completely conserved because the most recent `kick' steps did not
happen simultaneously for these bodies. Early in the simulation, we
find that this problem tends to trigger runaway growth at the timestep
boundaries first. This issue carries itself forward through the embryo
formation phase, and protoplanets tend to form at the boundaries. To
correct this issue, we ignore collisions between bodies on different
timesteps early in the simulation. We find that preventing
multi-timestep collisions until after the maximum mass grows by a
factor of 10 prevents any artifacts from developing at the timestep
boundaries, while also minimizing the number of `skipped' collisions. \textbf{In the case of the fdHi simulation, only about 20 collisions out of an eventual 900,000 are ignored. To verify that this timestepping scheme does not affect protoplanet growth, we tested an annulus of growing planetesimals with both fixed steps and our two-phase variable timestepping scheme. The results of these tests are shown in appendix \ref{sec:rung_egg}.}

@@

The timescales for embryo formation depend on
the chosen surface density profile, along with the local orbital
timescale. Protoplanets form first at the inner edge of the disk,
where the dynamical timescales are short. Growth proceeds in an
inside-out fashion, with the outermost regions of the disk completing
the protoplanet assembly phase last \textbf{((as an example, see figure 1 of \citet{kokubo02)}. This radial timescale dependence is not typically
accounted for in planet formation simulations\footnote{Instead, the innermost protoplanets 
patiently `wait' for this phase of evolution to complete everywhere in the disk} \textbf{(a notable exception being \citet{emsenhuber21a, emsenhuber21b})}, and appears to be an
important component to forming realistic solar system analogs
\citep{clement20}. As with the narrow annulus simulations, we stop the
integration once the masses of protoplanets in the outermost region of
the disk reach a steady value. In table \ref{tab:sim_properties}, we
summarize the outcomes of the four ``full disk'' cases.
